Tidy data structure to support exploration and modeling of temporal-context data

### 2018-06-14 main.Rmd
added explanations to tb data

# Introduction

@@

# Data semantics

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represent time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" works similarly as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database.

@@

```{r load-data}
```

(ref:tb-sub) A subset of estimates of tuberculosis burden generated by World Health Organisation in 2011 and 2012.

## Time index and interval

@@

For data indexed in regular time space, the time interval is obtained by computing the greatest common divisor based on the non-negative differences of a given time index, although the data may include implicit time gaps. This implies that all the units in the table must share a common interval. It is relatively easy to determine if the data is daily or sub-daily since most statistical packages have built-in time representations for date-times and dates, for example their corresponding classes `POSIXct` and `Date` in R. However these turn out inappropriate for weekly, monthly and quarterly data. For example, the smallest time unit of monthly data is months in a year, which is equispaced points in a year. However, using the first day of the month can range from 28 to 31 days, which is irregularly spaced. A genuine year-month representation should avoid associating days with its format. Displaying the year and month components suffice to signal its monthly aggregated values instead of observed values at a particular timestamp.

@@

## Key

@@

# Data pipeline

@@

## Literature review

>>>>

# Introduction {#sec:intro}

@@

# Data semantics {#sec:semantics}

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represent time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" works similarly as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database. Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. To be more specific, the key is passed via `id(country | continent, gender)`: a geographic hierarchy of country nested within continent, crossed with a demographic factor. Detailed discussions about index and key specification will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is a single measure in this case, it is sufficiently flexible to add other measured variables, for example, adding the corresponding population to calibrate the count.

@@

```{r tb-sub}
```

(ref:tb-sub) A subset of estimates of tuberculosis burden generated by World Health Organisation in 2011 and 2012, with 12 observations and 5 variables. The index refers to column `year`, the key to multiple columns: `contry` nested under `continent` crossed with `gender`, and the measured variable to column `count`.

## Time index and interval {#sec:index}

@@

For data indexed in regular time space, the time interval is obtained by computing the greatest common divisor based on the non-negative differences of a given time index, although the data may include implicit time gaps. This implies that all the units in the table must share a common interval. It is relatively easy to determine if the data is daily or sub-daily since most statistical packages have built-in time representations for date-times and dates, for example their corresponding classes `POSIXct` and `Date` in R. However these turn out inappropriate for weekly, monthly and quarterly data. For example, the smallest time unit of monthly data is months in a year, which is equispaced points in a year. However, using the first day of the month can range from 28 to 31 days, which is irregularly spaced. A genuine year-month representation should avoid associating days with its format. Displaying the year and month components suffice to signal its monthly aggregated values instead of observed values at a particular timestamp. (coercion)

@@

## Key {#sec:key}

@@

# Data pipeline {#sec:pipeline}

@@

## Literature review (take this heading out)
